---
title: "Tutorial on R Software"
subtitle: MUMS Undergraduate Workshop
author: 
   "Pulong Ma and Wenjia Wang, SAMSI"
date: "2/25/2019"
output:
  ioslides_presentation: default
  beamer_presentation: default
  slidy_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## What is R?
R is a free software environment for statistical computing and graphics:

* is a different implementation of S language;

* provides a wide variety of statistical and graphical techniques, and is highly extensible;

* is available as Free Software under the terms of the Free Sofware Foundations' GNU General Public License in source code form;

* runs on a wide variety of UNIX platforms and similar systems (including FreeBSD and Linux), Windows and MacOS;

* has powerful IDE (integrated development environment) such as RStudio.

 


## Install R 

1. Download the most recent version of R at <https://cran.r-project.org>.
2. Start the R program:
      + on Windows and MacOS, this will usually mean double-click on the R application
      + on UNIX-like systems, type "R" at a shell prompt.
3. As a first step with R, start with the R help brower by typing **help.start()** in the R command window. 

## Install RStudio
1. Go to RStudio's website <https://www.rstudio.com/products/rstudio/download/>.
2. Click on "Download RStudio Desktop"
3. Click on versions recommended for your system.

## R Session

### launch R 

### launch RStudio

## Operators in R
```{r operators 1, echo=TRUE}
## addition: +
3 + 2
## subtraction: -
3 - 2 
## multiplication: *
3 * 2
```

## Operators in R
```{r operators 2, echo=TRUE}
## division: /
3 / 2
## exponentiation: ^ or **
3^2
## greater than: >
3 > 2
```

## Operators in R
```{r operators 3, echo=TRUE}
## greater than or equal to: >=
3 >= 2
## exactly equal to: ==
3 == 2
## not equal to: !=
3 != 2
```


## Creating New Variables
Use the assignment operator **<-** or **=** to create new variables.
```{r assign, echo=TRUE}
x <- 1
print(x)
x = 2
print(x)
```

## Data Types
R has a wide variety of data types including

* scalars,

* vectors (numerical, character, logical),

* matrices,

* data frames,

* and lists.


## Scalar
```{r scalar, echo=TRUE}
num = 3
print(num)
print(typeof(num))

```

## Scalar
```{r scalar 2, echo=TRUE}
num = 3.14
num.int = as.integer(num)
print(num.int)
print(typeof(num.int))
```

## Vector
```{r vector 1, echo=TRUE}
x = 1:3
print(x)
y = c(4, 5, 6, 7)
y[1]  # subsetting 
y[-1] # subsetting
```  

## Vector
```{r vector 2, echo=TRUE}
y[c(1,4)]  # subsetting 
y[-c(1,4)] # subsetting
z = c(y[c(1,4)], y[-c(1,4)])
print(z)
```  

## Matrix
```{r matrix 1, echo=TRUE}
a = seq(1, 9, length.out=9)
A = matrix(a, nrow=3, ncol=3)
print(A)
print(typeof(A))
print(class(A))
```

## Matrix
```{r matrix 2, echo=TRUE}
# subsetting the first two elements in the first column
A[1:2, 1] 

A[1:2, c(1,2)] 


A[1:2, ]  # subsetting the first two rows
```

## Array
```{r array 1, echo=TRUE}
b = seq(1, 8, by=1)
B = array(data=b, dim=c(2,2,2))
print(B)
```

## Array
```{r array 2, echo=TRUE}
class(B)
B1 = B[ , , 1]

B2 = B[ , , 2]

C = cbind(B1, B2)
print(C)

```


## Data Frame
```{r df 1, echo=TRUE}
df = data.frame(x = c(1, 5, 6), y = c(2, 4, 3))
print(df)
print(class(df))
```


## Data Frame
```{r df 2, echo=TRUE}
print(df$x)
print(df$y)
```

## List
```{r list 1, echo=TRUE}
l1 = list(1, 2, 3)
print(l1)
class(l1)
```

## List
```{r list 2, echo=TRUE}
names(l1) <- c("a", "b", "c")
print(l1)

```

## List
```{r list 3, echo=TRUE}
l1[[1]]  ## subsetting the first element
l1$a  ## subsetting the element named a
l1$a = 4
print(l1$a)
```

## Linear Algebra
```{r LA 1, echo=TRUE}
A = matrix(c(2, 3, 1, 5), nrow=2, ncol=2)

## transpose 
t(A) 

## matrix addition
B = matrix(c(2, 2, 3, 5), nrow=2, ncol=2)
A + B


```
## Linear Algebra
```{r LA 3, echo=TRUE}
## matrix multiplication
A %*% B
### elementwise multiplication
A * B 
```

## Linear Algebra
```{r LA 2, echo=TRUE}
A = matrix(c(2, 1, 1, 5), nrow=2, ncol=2)
b = c(1,2)
## solve the system Ax = b
solve(A, b)
## compute cholesky decomposition
R = chol(A)
## use triangular solvers
backsolve(R, backsolve(R, b, transpose=TRUE))
```


## Importing Data
Let's load the buit-in R data "Orange" using **data()**.

```{r load data, echo=TRUE}
data("Orange")
head(Orange)
names(Orange)
```


## Importing Data
Let's save the data into local disk.

```{r load data 1, echo=TRUE}
Tree = Orange$Tree
age = Orange$age
circ = Orange$circumference

# save three vectors into the .RData format
save(Tree, age, circ, file="mydata.RData")

# save circumference into the .csv format
write.csv(circ, file="mydata.csv")

# load the csv file again
mycirc = read.csv(file="mydata.csv")
```

## Writing Your Own Function
```{r function 1, echo=TRUE}
mysquare <- function(x){
  y = x^2    ## main body of the function
  return(y)  ## return variable y
}

y1=mysquare(3)
print(y1)

## load functions stored in your local disk
source("mysquare.R")  
y2=mysquare(4)
print(y2)
```

## Packages
```{r package 1, echo=TRUE}
## Install from repository

install.packages(c("mvtnorm", "plotrix"))

## load packages
library(mvtnorm)
library(plotrix)
```





## Basic Plots
```{r, eval = FALSE, echo=TRUE}
## Use of "plot" function
x = 1:1000/10
y = sin(x)
print(head(x))
plot(x, y)
```

## Basic Plots
```{r}
## Use of "plot" function
x = 1:1000/10
y = sin(x)
print(head(x))
plot(x, y)
```

## Basic Plots
```{r, echo=TRUE}
## Use of "plot" function
plot(x, y, type = "l")
```

## Basic Plots
```{r, echo=TRUE}
plot(x[1:20], y[1:20], type = 'b', col = 'red',
     lwd = 5, ylim = c(0, 2))
```

## Basic Plots
```{r, eval=FALSE,echo=TRUE}
## Use of "hist" function
set.seed(2019)
data = rnorm(1000)
plot(data, type = 'l')
```

## Basic Plots
```{r}
## Use of "hist" function
set.seed(2019)
data = rnorm(1000)
plot(data, type = 'l')
```


## Basic Plots
```{r, echo=TRUE}
## Use of "hist" function
hist(data)
```

## Basic Plots
```{r, echo=TRUE}
## Use of "hist" function
hist(data, breaks = 30)
```

## Sampling from Basic Distributions


```{r, echo=TRUE}
## Uniform distribution
runif(n = 2, min = 0, max = 1)

## Normal distribution 
rnorm(n = 2, mean = 0, sd = 1)
```


## Multivariate Normal Distribution
```{r}
options(warn=-1)
library("mvtnorm")
library("plotrix")
```

```{r, echo=TRUE}
## Need package mvtnorm
Sigma <- matrix(c(10,3,3,2),2,2)
rmvnorm(n = 1, mean = rep(0, nrow(Sigma)), 
        sigma = Sigma)

```

## Calculating $\pi$
```{r, eval = FALSE,echo = TRUE}
## Need package plotrix
plot(0.5,0.5,type ="p",asp = 1, xlim=c(0,1)
     ,ylim=c(0,1),color = "black")

draw.circle(0.5,0.5,0.5,nv=1000
            ,border=NULL,col=NA,lty=1,lwd=1)

rect( 0, 0, 1, 1) 
```

## Calculating $\pi$
```{r}
## Need package plotrix
plot(0.5,0.5,type ="p",asp = 1, xlim=c(0,1)
     ,ylim=c(0,1),color = "black")

draw.circle(0.5,0.5,0.5,nv=1000
            ,border=NULL,col=NA,lty=1,lwd=1)

rect( 0, 0, 1, 1) 

```

## Calculating $\pi$
```{r}
plot.new()
plot(0.5,0.5,type ="p",asp = 1, xlim=c(0,1),ylim=c(0,1),color = "black")
draw.circle(0.5,0.5,0.5,nv=1000,border=NULL,col=NA,lty=1,lwd=1)
rect( 0, 0, 1, 1) 
mcpiwplot <- function(n){
  m = 0
  x <- runif(n)
  y <- runif(n)
  x1 <- x - 0.5
  y1 <- y - 0.5 #circle has center at (0.5,0.5)
  r2 <- x1^2 + y1^2
  for(i in 1:n){
    if(r2[i] <= 0.25){
      m <- m+1
      points(x[i],y[i] , col = "red")
    }
    else{
      points(x[i],y[i], col = "blue")
    }
  }
  piapprox <- m/(0.25*n)
  return(((piapprox-pi)/pi))
}
mcpi <- function(n){
  m = 0
  x <- runif(n)
  y <- runif(n)
  x1 <- x - 0.5
  y1 <- y - 0.5 #circle has center at (0.5,0.5)
  r2 <- x1^2 + y1^2
  for(i in 1:n){
    if(r2[i] <= 0.25){
      m <- m+1
    }
  }
  piapprox <- m/(0.25*n)
  return(((piapprox-pi)/pi))
}
mcpiwplot(50)
```

## Convergence Rate
```{r}
m = 20
error <- 1:m
for(i in 1:m){
  errore <- 1:100
  for(j in 1:100){
    errore[j] <- mcpi(i*15)
  }
  error[i] <- sum(abs(errore))/100
}
plot.new()
plot(1:m*20,error,type = "l",xlab = "Sample size", ylab ="Error" )

```

